The \project{unRAVE} catalog of stellar atmospheric parameters and\\ chemical abundances for chemo-dynamic studies in the \gaia\ era

### 2016-09-09 article/unrave.tex
Update figures to 0.97 to incl error floors; fix minor things

\author{
Andrew R. Casey\altaffilmark{1}, 
Keith Hawkins\altaffilmark{1}, 
David W. Hogg\altaffilmark{2,3,4,5}, 
Melissa Ness\altaffilmark{5}, 
Hans Walter-Rix\altaffilmark{5}, 
Georges Kordopatis\altaffilmark{6}, 
Andrea Kunder\altaffilmark{6}, 
Matthias Steinmetz\altaffilmark{6}, 
Sergey Koposov\altaffilmark{1}, 
Harry Enke\altaffilmark{6}, 
Jason Sanders\altaffilmark{1}, 
Gerry Gilmore\altaffilmark{1}, 
Toma\v{z} Zwitter\altaffilmark{6}, 
Kenneth C. Freeman\altaffilmark{7}, 
Luca Casagrande\altaffilmark{7}, 
Gal~Matijevic\altaffilmark{6}, 
George Seabroke\altaffilmark{8}, 
Olivier Bienayme\altaffilmark{9}, 
Joss Bland-Hawthorn\altaffilmark{10}, 
Brad~K.~Gibson\altaffilmark{11}, 
Eva~K.~Grebel\altaffilmark{12}, 
Amina Helmi\altaffilmark{13}, 
Ulisse Munari\altaffilmark{14}, 
Julio F. Navarro\altaffilmark{15,16}, 
Warren Reid\altaffilmark{17,18}, 
Arnaud Siebert\altaffilmark{19},
Rosemary Wyse\altaffilmark{20}\\
\vspace{2em}
\emph{(Affiliations can be found at the end of the article)}
}

\begin{abstract}
The Milky Way is a powerful laboratory for understanding galaxy formation,
as it can provide orbits, ages, atmospheric parameters, as well as chemical 
abundances for vast sets of individual stars.  These inferences require 
both astrometric and spectroscopic data.  Indeed, in order to be maximally
useful for chemo-dynamic studies, the Tycho-Gaia Astrometric Solution 
(\tgas) sample in the first \gaia\ data release requires a spectroscopic 
complement that includes radial velocities, stellar atmospheric parameters, and 
elemental abundances.  Among existing spectroscopic samples, the RAdial 
Velocity Experiment (\rave) survey has the largest overlap with \tgas: 
up to 264,276 stars.  We present a data-driven re-analysis of \rave\ 
spectra using \thecannon, and we provide more precise and accurate stellar
parameters and abundances than available in previous \rave\ data releases. 
We constructed our model using high-fidelity \apogee\ stellar parameters 
and abundances for red giant branch stars that overlap with \rave, and 
stellar parameters from the \project{K2}/\project{Ecliptic Pole Input 
Catalog} for main-sequence and sub-giant branch stars.  We derive, and 
validate, improved effective temperature $\teff$, surface gravity $\logg$, 
and chemical abundances of up to seven elements (O, Mg, Al, Si, Ca, Fe, Ni) 
for \Nstarsqc\ stars.  In total we report \Nabundances\ elemental 
abundances, with a typical precision of 0.07~dex.  The \unrave\ catalog 
presented here, in conjunction with \tgas, represents the most powerful 
data set for chemo-dynamic analysis at the dawn of broadly available 
\gaia\ data.
\end{abstract}

@@

We inferred labels for all \Nspectra\ \rave\ spectra using this main-sequence/sub-giant
star model; we made no \emph{a priori} decisions as to whether a star was likely a
main-sequence/sub-giant star or not.  The results for the entire survey sample are shown
in the centre panel of Figure \ref{fig:test-set-density}.  The increased density of 
solar-type stars is consistent with \rave\ observing stars in the local neighbourhood, 
and the high number of turn-off and main-sequence stars relative to the sub-giant branch 
is expected from the relative lifetimes of these evolutionary phases.  An over-density
of stars near the base of the giant branch is also present.  This artefact is due 
to having giant stars in the test set, but not in the training set, and the model is 
(poorly) extrapolating outside the convex hull of the training set.

@@

Using positions, proper motions, and metallicities from the \rave\ official
data release (such that they can be used as comparison), we identify $\sim$160 
probable members of four open clusters that were observed by \rave.
Specifically we identify 78 potential Pleiades members, 26 candidates in the
Hyades, another 13 in IC4561, and 30 stars in the solar-metallicity open
cluster M67.
%Upon requiring $(\alpha,\delta) = (51.5^\circ$ to $57.5^\circ$, 21.5$^\circ$ to 28$^\circ)$,
%have error in metallicity \texttt{e\_M\_H\_K} $< 0.5$~dex, and radial
%velocities between $-10$~km~s$^{-1}$ and 20~km~s$^{-1}$, we identify 78 
%Pleiades members.  For the Hyades, by requiring stars within $3.5^\circ$ of the
%cluster centre to have: radial velocities between 36~km~s$^{-1}$ and 44~km~s$^{-1}$,
%$[{\rm M/H}]_{RAVE}$ within $-0.15$ and 0.35, and \rave\ metallicity errors
%$<0.12$~dex, we identify 26 potential members.  Similarly for IC4561 we select
%13 stars by requiring: radial velocities within $-20$~km~s$^{-1}$ to $-40$~km~s$^{-1}$,
%an error in radial velocity $<3$~km~s$^{-1}$; $[{\rm M/H}]_{RAVE}$ within $-0.15$ and 0.35;
%$\alpha$ between $261.1^\circ$ and $261.6^\circ$; and $\delta$ within $-49.7^\circ$
%and $-50.1^\circ$.  For the solar-metallicity open cluster member, we 30 select stars solely on
%position: $(\alpha,\delta) = (133.1^\circ$ to $132.5^\circ$, $11.6^\circ$ to $12^\circ)$.
We show the effective temperature $\teff$ and surface gravity $\logg$ for these
cluster candidates in Figure \ref{fig:open-cluster-HRD}.  The isochrones are 
sourced from \citet{Bressan_2012}, with cluster properties adopted from 
\cite{Kharchenko_2013}.

@@

For practical purposes we adopted separate models: one for the giant branch and one
for the main-sequence.  A third model was used to derive relative weights for which
results to use.  The relative weighting we have used does not have any formal
interpretation as a likelihood or belief (in any sense): it was introduced for
practical reasons to identify systematic errors and combine results for multiple
models.  Because the relative weights have no formal interpretation, it is reasonable
to consider this method is as \emph{ad hoc} as any other approach.  The relative 
weighting has no warranty to be (formally) correct, and may introduce inconsistencies
or systematic errors rather than minimizing them.

If we only consider the results from individual models, there are a number of cautionary
remarks that stem from the construction of the training set.  The labels for red giant 
branch stars primarily come from \apogee, where previous successes with \thecannon\
have demonstrated that \apogee\ labels based on high S/N data can be of high fidelity
\citep{Ness_2015,Ness_2016,Ho_2016,Casey_2016b}.
However, the lack of metal-poor stars in the \apogee/\rave\ overlap sample produced
a tapering-off in the test set --- where \emph{no} stars had reliably reported metallicities
below that of the training set --- which forced us to construct a heterogeneous training
set.  The metal-poor stars included in this sample are from high-resolution studies
\citep{Fulbright_2010,Ruchti_2011}, but it is not known if the stellar parameters are
of high fidelity because we have limited quality metrics available.  Moreover, there
is no guarantee that the stellar parameters \emph{or} abundances are on the same scale
as \apogee\ \citep[and good reasons to believe they will not be; see][]{Smiljanic_2014}.

@@

This choice is reflected in the abundances of the test set.  While we do recover
trustworthy metallicities for ultra metal-poor ($[{\rm Fe/H}] \lesssim -4$) stars like 
CD-38~245, the individual abundances for all extremely metal-poor stars aggregate
(in [X/Fe] space; Figure~\ref{fig:gce}) at the assumed abundances for the metal-poor
stars in our sample.  Thus, while the overall metallicities appear reliable, the 
individual abundances for extremely metal-poor stars in the test set cannot be
considered trustworthy in any sense.  For this reason we have updated the electronic
catalog to discard these results, and included a quality control flag to indicate
the result was discarded (see catalog \texttt{README} file for specifics related to
the data structure and content).

@@

We have shown some potential outcomes when the test set spectra differ 
significantly from the spectra in the training set.  Test set spectra that is
`unusual' from the training set can be projected as peculiar artefacts in 
label space.  In other words, unusual spectra can appear as `clumps' in
regions of parameter space that we could consider as being normal (e.g.,
an over-density of solar-type stars).  We addressed this issue for the main-sequence
and giant models by using a third model (Section \ref{sec:a-simple-model}) to
calculate relative weights.  However, spectra that are unusual from the 
training set used in the simple model still project as systematic artefacts.

@@

We have included a boolean flag column in our catalog for quality control,
to signify whether a result is thought to be reasonable.  Specifically, 
stars where \texttt{SPARV} indicates a $\teff > 8000$~K do not meet our quality 
control flag.  For all other stars, we further require $\chi_r^2 < 3$ and
$S/N >10$~pixel$^{-1}$.  These quality constraints represent the minimum
acceptable constraints and should not be taken verbatim: judicious use of
the morphological classifications \citep{Matijevic_2012} or additional 
metadata from the \rave\ pre-processing pipelines is strongly encouraged.

@@

\begin{figure*}[p]
\includegraphics[width=\textwidth]{figures/pairwise-metrics.pdf}
\caption{Distribution of differences in label estimates from multiple visits, divided by the quadrature sum of their formal errors, and a minimum error value for each observation.  If our measurements were unbiased and our errors were representative, no minimum error floor would be required and this distribution would by normally-distributed with zero mean and unity variance. 
We increased the error floor for a given label until the distribution variance approximately reached unity. The error floors are given for each label.\label{fig:pairwise-comparison}}
\end{figure*}

@@

\begin{figure}[p]
\center
\includegraphics[width=0.65\textwidth]{figures/hot-stars.pdf}
\caption{Effective temperatures $\teff$ and surface gravities $\logg$ from this work, where each hexagonal bin is colored by the \emph{maximum} temperature for any star in that bin, as reported by the \rave\ pre-processing pipeline \texttt{SPARV} \citep{Steinmetz_2006,Zwitter_2008}.  Hot stars are not included in our training set, and appear above the turnoff in our labels.  For this reason, we supply a boolean quality control flag \texttt{QC} that flags stars with $\teff > 8000$~K (as reported by \texttt{SPARV}), and other results that are of poor quality.\label{fig:hot-stars}}
\end{figure}

>>>>

\author{
Andrew R. Casey\altaffilmark{1}, 
Keith Hawkins\altaffilmark{1}, 
David W. Hogg\altaffilmark{2,3,4,5}, 
Melissa Ness\altaffilmark{5}, 
Hans Walter-Rix\altaffilmark{5}, 
Georges Kordopatis\altaffilmark{6}, 
Andrea Kunder\altaffilmark{6}, 
Matthias Steinmetz\altaffilmark{6}, 
Sergey Koposov\altaffilmark{1}, 
Harry Enke\altaffilmark{6}, 
Jason Sanders\altaffilmark{1}, 
Gerry Gilmore\altaffilmark{1}, 
Toma\v{z} Zwitter\altaffilmark{6}, 
Kenneth C. Freeman\altaffilmark{7}, 
Luca Casagrande\altaffilmark{7}, 
Gal~Matijevi\v{c}\altaffilmark{6}, 
George Seabroke\altaffilmark{8}, 
Olivier Bienayme\altaffilmark{9}, 
Joss Bland-Hawthorn\altaffilmark{10}, 
Brad~K.~Gibson\altaffilmark{11}, 
Eva~K.~Grebel\altaffilmark{12}, 
Amina Helmi\altaffilmark{13}, 
Ulisse Munari\altaffilmark{14}, 
Julio F. Navarro\altaffilmark{15,16}, 
Warren Reid\altaffilmark{17,18}, 
Arnaud Siebert\altaffilmark{19},
Rosemary Wyse\altaffilmark{20}\\
\vspace{2em}
\emph{(Affiliations can be found at the end of the article)}
}

\begin{abstract}
The Milky Way is a powerful laboratory for understanding galaxy formation,
as it can provide orbits, ages, atmospheric parameters, as well as chemical 
abundances for vast sets of individual stars.  These inferences require 
both astrometric and spectroscopic data.  Indeed, in order to be maximally
useful for chemo-dynamic studies, the Tycho-Gaia Astrometric Solution 
(\tgas) sample in the first \gaia\ data release requires a spectroscopic 
complement that includes radial velocities, stellar atmospheric parameters, and 
elemental abundances.  Among existing spectroscopic samples, the RAdial 
Velocity Experiment (\rave) survey has the largest overlap with \tgas: 
over 200,000 stars.  We present a data-driven re-analysis of \rave\ 
spectra using \thecannon, and we provide more precise and accurate stellar
parameters and abundances than available in previous \rave\ data releases. 
We constructed our model using high-fidelity \apogee\ stellar parameters 
and abundances for red giant branch stars that overlap with \rave, and 
stellar parameters from the \project{K2}/\project{Ecliptic Pole Input 
Catalog} for main-sequence and sub-giant branch stars.  We derive, and 
validate, improved effective temperature $\teff$, surface gravity $\logg$, 
and chemical abundances of up to seven elements (O, Mg, Al, Si, Ca, Fe, Ni) 
for \Nstarsqc\ stars.  In total we report \Nabundances\ elemental 
abundances, with a typical precision of 0.07~dex.  The \unrave\ catalog 
presented here, in conjunction with \tgas, represents the most powerful 
data set for chemo-dynamic analysis at the dawn of broadly available 
\gaia\ data.
\end{abstract}

@@

We inferred labels for all \Nspectra\ \rave\ spectra using this main-sequence/sub-giant
star model; we made no \emph{a priori} decisions as to whether a star was likely a
main-sequence/sub-giant star or not.  The results for the entire survey sample are shown
in the center panel of Figure \ref{fig:test-set-density}.  The increased density of 
solar-type stars is consistent with \rave\ observing stars in the local neighbourhood, 
and the high number of turn-off and main-sequence stars relative to the sub-giant branch 
is expected from the relative lifetimes of these evolutionary phases.  An over-density
of stars near the base of the giant branch is also present.  This artefact is due 
to having giant stars in the test set, but not in the training set, and the model is 
(poorly) extrapolating outside the convex hull of the training set.

@@

Using positions, proper motions, and metallicities from the \rave\ survey
(i.e., not ours, such that they can be used as comparison), we identify $\sim$160 
probable members of four open clusters that were observed by \rave.
Specifically we identify 78 potential Pleiades members, 26 candidates in the
Hyades, another 13 in IC4561, and 30 stars in the solar-metallicity open
cluster M67.
%Upon requiring $(\alpha,\delta) = (51.5^\circ$ to $57.5^\circ$, 21.5$^\circ$ to 28$^\circ)$,
%have error in metallicity \texttt{e\_M\_H\_K} $< 0.5$~dex, and radial
%velocities between $-10$~km~s$^{-1}$ and 20~km~s$^{-1}$, we identify 78 
%Pleiades members.  For the Hyades, by requiring stars within $3.5^\circ$ of the
%cluster center to have: radial velocities between 36~km~s$^{-1}$ and 44~km~s$^{-1}$,
%$[{\rm M/H}]_{RAVE}$ within $-0.15$ and 0.35, and \rave\ metallicity errors
%$<0.12$~dex, we identify 26 potential members.  Similarly for IC4561 we select
%13 stars by requiring: radial velocities within $-20$~km~s$^{-1}$ to $-40$~km~s$^{-1}$,
%an error in radial velocity $<3$~km~s$^{-1}$; $[{\rm M/H}]_{RAVE}$ within $-0.15$ and 0.35;
%$\alpha$ between $261.1^\circ$ and $261.6^\circ$; and $\delta$ within $-49.7^\circ$
%and $-50.1^\circ$.  For the solar-metallicity open cluster member, we 30 select stars solely on
%position: $(\alpha,\delta) = (133.1^\circ$ to $132.5^\circ$, $11.6^\circ$ to $12^\circ)$.
We show the effective temperature $\teff$ and surface gravity $\logg$ for these
cluster candidates in Figure \ref{fig:open-cluster-HRD}.  The isochrones are 
sourced from \citet{Bressan_2012}, with cluster properties adopted from 
\cite{Kharchenko_2013}.

@@

For practical purposes we adopted separate models: one for the giant branch and one
for the main-sequence.  A third model was used to derive relative weights for which
results to use.  The relative weighting we have used does not have any formal
interpretation as a likelihood or belief (in any sense): it was introduced for
practical reasons to identify systematic errors and combine results for multiple
models.  Because the relative weights have no formal interpretation, it is reasonable
to consider this method is as \emph{ad hoc} as any other approach.  The relative 
weighting has no warranty to be (formally) correct, and therefore may introduce 
inconsistencies or systematic errors rather than minimizing them.

If we only consider the results from individual models, there are a number of cautionary
remarks that stem from the construction of the training set.  The labels for red giant 
branch stars primarily come from \apogee, where previous successes with \thecannon\
have demonstrated that \apogee\ labels based on high S/N data can be of high fidelity
\citep{Ness_2015,Ness_2016,Ho_2016,Casey_2016b}.
However, the lack of metal-poor stars in the \apogee/\rave\ overlap sample produced
a tapering-off in the test set --- where \emph{no} stars had reliably reported metallicities
below that of the training set --- which forced us to construct a heterogeneous training
set.  The metal-poor stars included in this sample are from high-resolution studies
\citep{Fulbright_2010,Ruchti_2011}, but it is not known if the stellar parameters are
of high fidelity because we have a limited number of quality statistics available. 
Moreover, there is no guarantee that the stellar parameters \emph{or} abundances are 
on the same scale as \apogee\ \citep[and good reasons to believe they will not be; see][]{Smiljanic_2014}.

@@

This choice is reflected in the abundances of the test set.  While we do recover
trustworthy metallicities for ultra metal-poor ($[{\rm Fe/H}] \lesssim -4$) stars like 
CD-38~245, the individual abundances for all extremely metal-poor stars aggregate
(in [X/Fe] space; Figure~\ref{fig:gce}) at the assumed abundances for the metal-poor
stars in our sample.  Thus, while the overall metallicities appear reliable, the 
individual abundances for extremely metal-poor stars in the test set cannot be
considered trustworthy in any sense.  For this reason we have updated the electronic
catalog to discard these results as erroneous.

@@

We have shown some potential outcomes when the test set spectra differ 
significantly from the spectra in the training set.  Test set spectra that is
`unusual' from the training set can be projected as peculiar artefacts in 
label space.  In other words, unusual spectra can appear as `clumps' in
regions of parameter space that we could consider as being normal (e.g.,
an over-density of solar-type stars).  We addressed this issue for the main-sequence
and giant models by using a third model (Section \ref{sec:a-simple-model}) to
calculate relative weights.  However, spectra that are unusual from the 
training set used in the simple model could still project as systematic 
artefacts in label space.

@@

This line of reasoning extends to spectra with other peculiar characteristics 
(e.g., chromospheric emission), and for these reasons we recommend the use of
additional metadata to investigate possible artefacts.  In our catalog we have
included a column containing a boolean flag to indicate whether the labels pass
very weak quality constraints.  Specifically, we flag results as failing our
quality constraints if \texttt{SPARV} indicates a $\teff > 8000$~K, or if 
$\chi_r^2 > 3$, or if $S/N < 10$~pixel$^{-1}$.  These quality constraints 
represent the minimum acceptable conditions and should not be taken verbatim: 
judicious use of the morphological classifications \citep{Matijevic_2012} or 
additional metadata from the \rave\ pre-processing pipelines is strongly 
encouraged.

@@

\begin{figure*}[p]
\includegraphics[width=\textwidth]{figures/pairwise-metrics.pdf}
\caption{Distribution of differences in label estimates from multiple visits, divided by the quadrature sum of their formal errors, and a minimum error value for each observation.  If our measurements were unbiased and our errors were representative, no minimum error floor would be required and the distribution of $\eta$ would by normally-distributed with zero mean and unity variance. 
We increased the error floor for each label until the variance in the distribution of $\eta$ approximately reached unity. Derived error floors are shown for each label.\label{fig:pairwise-comparison}}
\end{figure*}

@@

\begin{figure}[p]
\center
\includegraphics[width=0.65\textwidth]{figures/hot-stars.pdf}
\caption{Effective temperatures $\teff$ and surface gravities $\logg$ from this work, where each hexagonal bin is colored by the \emph{maximum} temperature for any star in that bin, as reported by the \rave\ pre-processing pipeline \texttt{SPARV} \citep{Steinmetz_2006,Zwitter_2008}.  Hot stars are not included in our training set, and appear above the turnoff in our labels.  For this reason, we supply a boolean quality control flag \texttt{QC} that flags stars (as \texttt{False}, for failing to meet our quality constraints) with $\teff > 8000$~K (as reported by \texttt{SPARV}), or results that do not meet other minimum constraints (see Section \ref{sec:discussion} for details).\label{fig:hot-stars}}
\end{figure}
